\section*{List of publications included in the thesis}
\addcontentsline{toc}{section}{List of publications included in the thesis}
\mark{{List of publications included in the thesis}}
% Add entry in the table of contents as section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This thesis is based on the following papers:

\subsubsection{Chapter~\ref{chapter:statical-challenges-2021}}
\vspace{-1em}
\noindent\underline{J. Malato}, L. Graça, L. Nacul, E. M. Lacerda, and N. Sepúlveda. Statistical challenges of investigating a disease with a complex diagnosis. In P. Milheiro, A. Pacheco, B. de Sousa, I. F. Alves, I. Pereira, M. J. Polidoro, and S. Ramos, editors, \textit{Estatística: Desafios transversais às ciências com dados}, Eds. Sociedade Portuguesa de Estatística, pages 153--167. Sociedade Portuguesa de Estatística, 2021. ISBN 978-972-8890-47-6. URL \url{https://www.spestatistica.pt/storage/app/uploads/public/609/28f/6d0/60928f6d08a0c016386627.pdf}.

\subsubsection{Chapter~\ref{chapter:misdiagnosis-serology-2022}}
\vspace{-1em}
\noindent\underline{J. Malato}, L. Graça, and N. Sepúlveda. Impact of misclassification and imperfect serological tests in association analyses of ME/CFS applied to COVID-19 data. In R. Bispo, L. Henriques-Rodrigues, R. Alpizar-Jara, and M. de Carvalho, editors, \textit{Recent Developments in Statistics and Data Science}, Springer Proceedings in Mathematics \& Statistics, pages 215--225, Cham, 2022. Springer International Publishing. ISBN 978-3-031-12766-3. doi: 10.1007/978-3-031-12766-315.

\subsubsection{Chapter~\ref{chapter:impact-misdignosis-2023}}
\vspace{-1em}
\noindent\underline{J. Malato}, L. Graça, and N. Sepúlveda. Impact of misdiagnosis in case-control studies of myalgic encephalomyelitis/chronic fatigue syndrome. \textit{Diagnostics}. 2023; 13(3):531. doi: \url{https://doi.org/10.3390/diagnostics13030531}.

\subsubsection{Chapter~\ref{chapter:2022-revisiting-igg}}
\vspace{-1em}
\noindent N. Sepúlveda, \underline{J. Malato}, F. Sotzny, A.D. Grabowska, A. Fonseca, C. Cordeiro, L. Graça, P. Biecek, U. Behrends, J. Mautner, F. Westermeier, E.M. Lacerda, and C. Scheibenbogen. Revisiting IgG Antibody Reactivity to Epstein-Barr Virus in Myalgic Encephalomyelitis/Chronic Fatigue Syndrome and Its Potential Application to Disease Diagnosis. \textit{Frontiers in Medicine}. 2022. 9:921101. doi: \url{https://doi.org/10.3389/fmed.2022.921101}.

\subsubsection{Chapter~\ref{chapter:2023-sym-and-herpesvirus}}
\vspace{-1em}
\noindent T.D. Domingues, \underline{J. Malato}, A.D. Grabowska, J.-S. Lee, J. Ameijeiras-Alonso, P. Biecek, L. Graça, H. Mouriño, C. Scheibenbogen, F. Westermeier, L. Nacul, J.M. Cliff, E. Lacerda, and N. Sepúlveda. Association analysis between symptomology and herpesvirus IgG antibody concentrations in myalgic encephalomyelitis/chronic fatigue syndrome (ME/CFS) and multiple sclerosis. \textit{Heliyon}. 2023; 9(7):E18250. doi: \url{https://doi.org/10.1016/j.heliyon.2023.e18250}.

\subsubsection{Chapter~\ref{chapter:2021-ace-ace2}}
\vspace{-1em}
\noindent\underline{J. Malato}, F. Sotzny, S. Bauer, H. Freitag, A. Fonseca, A.D. Grabowska, L. Graça, C. Cordeiro, L. Nacul, E.M. Lacerda, J. Castro-Marrero, C. Scheibenbogen, F. Westermeier, and N. Sepúlveda. The SARS-CoV-2 receptor angiotensin-converting enzyme 2 (ACE2) in myalgic encephalomyelitis/chronic fatigue syndrome: A meta-analysis of public DNA methylation and gene expression data. \textit{Heliyon}. 2021; 7(8). doi: \url{https://doi.org/10.1016/j.heliyon.2021.e07665}.

\subsubsection{Chapter~\ref{chapter:2022-covid19-01}}
\vspace{-1em}
\noindent\underline{J. Malato}, R.M. Ribeiro, P.P. Leite, P. Casaca, E. Fernandes, C. Antunes, V.R. Fonseca, M.C. Gomes, and L. Graca. Risk of BA.5 infection among persons exposed to previous SARS-CoV-2 variants. \textit{New England Journal of Medicine}. 2022; 387(10):953--954. doi: \url{https://doi.org/10.1056/NEJMc2209479}.

\subsubsection{Chapter~\ref{chapter:2023-covid19-02}}
\vspace{-1em}
\noindent\underline{J. Malato}, R.M. Ribeiro, E. Fernandes, P.P. Leite, P. Casaca, C. Antunes, V.R. Fonseca, M.C. Gomes, and L. Graca. Stability of hybrid versus vaccine immunity against BA.5 infection over 8 months. \textit{The Lancet Infectious Diseases}. 2023; 23(2):148--150. doi: \url{https://doi.org/10.1016/S1473-3099(22)00833-7}.

\vfill
